---
title: "Automated Parameter Estimation for Camera Trap Density Models"
subtitle: "A worked example using the CV-enhanced distance sampling pipeline"
author: "Sierra McMurry, Ben Goldstein, Mohammed Alyetam, Roland Kays"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# Introduction

This vignette demonstrates the complete parameter estimation pipeline described in McMurry et al. (in review). Starting from AI-derived animal coordinates and Wildlife Insights metadata, we estimate all four parameters required for camera trap density models:

1. **Movement speed** — via Size-Biased Distribution (SBD) correction
2. **Activity level** — via circular kernel density estimation
3. **Staying time** — via bounding box interpolation
4. **Effective detection distance (EDD)** — via a joint multi-species hierarchical model

These parameters are direct inputs for the Random Encounter Model (REM; Rowcliffe et al. 2008), Camera Trap Distance Sampling (CTDS; Howe et al. 2017), and the Random Encounter and Staying Time model (REST; Nakashima et al. 2018).

## Example dataset

We use 28 camera deployments from a montane forest array in Montana, USA, containing approximately 20,000 AI-processed frames across 9 mammal species.

## Prerequisites

```{r packages}
library(tidyverse)
library(activity)    # Activity level estimation
library(nimble)      # Joint EDD model
library(MCMCvis)     # MCMC diagnostics
library(coda)        # MCMC convergence
```

# 1. Load and inspect data

## 1.1 Input files

The pipeline requires two inputs:

1. **AI coordinate output** — a CSV with one row per detected animal per frame, containing world coordinates (world_x, world_z) estimated by the depth pipeline, bounding box coordinates, timestamps, and species identifications.
2. **Deployment metadata** — start/end dates, camera model, and location for each deployment.

```{r load-data}
# Load the example dataset
data <- read.csv("example_vignette_data.csv", stringsAsFactors = FALSE)
data$timestamp_clean <- as.POSIXct(data$timestamp_clean, tz = "UTC")

cat("Frames:", nrow(data), "\n")
cat("Species:", n_distinct(data$common_name_clean), "\n")
cat("Deployments:", n_distinct(data$deployment_id_clean), "\n")
```

## 1.2 Inspect the data structure

Key columns from the AI pipeline:

```{r inspect}
data %>%
  select(deployment_id_clean, sequence_id_use, common_name_clean,
         timestamp_clean, world_x, world_z, x1, y1, x2, y2,
         detection_confidence) %>%
  head(10)
```

- `world_x`: lateral position (meters) relative to camera
- `world_z`: forward distance (meters) from camera (used as detection distance)
- `x1, y1, x2, y2`: bounding box pixel coordinates
- `sequence_id_use`: groups frames from the same detection event

## 1.3 Species summary

```{r species-summary}
data %>%
  group_by(common_name_clean) %>%
  summarise(
    n_frames = n(),
    n_sequences = n_distinct(sequence_id_use),
    n_deployments = n_distinct(deployment_id_clean),
    .groups = "drop"
  ) %>%
  arrange(desc(n_frames)) %>%
  knitr::kable(col.names = c("Species", "Frames", "Sequences", "Deployments"))
```

# 2. Quality filtering

Before estimating parameters, we apply quality filters to remove unreliable AI coordinate estimates. Filters are applied at two levels: frame-level (for distance estimation) and sequence-level (for speed estimation).

## 2.1 Frame-level filters

```{r frame-filters}
# Get image dimensions for bounding box area calculation
# Assuming standard camera trap resolution; adjust for your cameras
IMG_WIDTH <- 2048
IMG_HEIGHT <- 1536

data <- data %>%
  mutate(
    # Bounding box area as proportion of frame
    bbox_width = x2 - x1,
    bbox_height = y2 - y1,
    bbox_area_pct = (bbox_width * bbox_height) / (IMG_WIDTH * IMG_HEIGHT) * 100,
    
    # Frame-level flags
    flag_bbox_too_large = bbox_area_pct > 80,
    flag_too_close = world_z < 1,
    flag_too_far = world_z > 26,
    flag_any_frame = flag_bbox_too_large | flag_too_close | flag_too_far
  )

# Summary of frame-level filtering
cat("Frame-level filter summary:\n")
cat("  Bbox >80% of frame:", sum(data$flag_bbox_too_large, na.rm = TRUE), "\n")
cat("  Too close (<1m):", sum(data$flag_too_close, na.rm = TRUE), "\n")
cat("  Too far (>26m):", sum(data$flag_too_far, na.rm = TRUE), "\n")
cat("  Total flagged:", sum(data$flag_any_frame, na.rm = TRUE),
    "(", round(mean(data$flag_any_frame, na.rm = TRUE) * 100, 1), "%)\n")

# Remove flagged frames for distance estimation
data_clean <- data %>% filter(!flag_any_frame)
cat("  Retained:", nrow(data_clean), "frames\n")
```

## 2.2 Add time variables

```{r time-vars}
data_clean <- data_clean %>%
  mutate(
    hour = as.numeric(format(timestamp_clean, "%H")) +
           as.numeric(format(timestamp_clean, "%M")) / 60 +
           as.numeric(format(timestamp_clean, "%S")) / 3600
  )
```

# 3. Movement speed estimation

Movement speed is estimated using the Size-Biased Distribution (SBD) framework (Rowcliffe et al. 2016). Camera traps over-sample faster-moving animals because they cross the detection zone more frequently. The SBD correction uses the harmonic mean to recover the true population mean speed.

## 3.1 Timestamp collapse

Camera traps operating in burst mode capture multiple frames within the same second. Small AI coordinate shifts between burst images accumulate as artificial movement. We collapse each sequence to one coordinate per unique timestamp.

```{r timestamp-collapse}
collapsed <- data_clean %>%
  group_by(sequence_id_use, deployment_id_clean, common_name_clean) %>%
  arrange(timestamp_clean) %>%
  mutate(
    # Unique second within sequence
    time_sec = as.numeric(difftime(timestamp_clean, min(timestamp_clean), units = "secs")),
    unique_sec = floor(time_sec)
  ) %>%
  # Keep first frame per unique second
  group_by(sequence_id_use, deployment_id_clean, common_name_clean, unique_sec) %>%
  slice(1) %>%
  ungroup()

cat("Before collapse:", nrow(data_clean), "frames\n")
cat("After collapse:", nrow(collapsed), "frames\n")
```

## 3.2 Compute frame-to-frame step distances

```{r step-distances}
collapsed <- collapsed %>%
  group_by(sequence_id_use) %>%
  arrange(timestamp_clean) %>%
  mutate(
    dx = world_x - lag(world_x),
    dz = world_z - lag(world_z),
    dt = as.numeric(difftime(timestamp_clean, lag(timestamp_clean), units = "secs")),
    step_distance = sqrt(dx^2 + dz^2),
    frame_speed = ifelse(dt > 0, step_distance / dt, NA_real_)
  ) %>%
  ungroup()
```

## 3.3 Coordinate anchoring

The AI-estimated position of a stationary animal can shift slightly due to posture changes (coordinate jitter). We use species-specific anchoring thresholds to zero out sub-threshold movements.

```{r anchoring}
# Species-specific anchoring thresholds (from Kneedle algorithm + manual overrides)
anchoring_thresholds <- tibble(
  common_name_clean = c("White-tailed Deer", "American Black Bear", "Mule Deer",
                         "Elk", "Snowshoe Hare", "Red Squirrel", "Bobcat",
                         "Coyote", "Moose", "Puma", "Wild Turkey", "Red Fox"),
  threshold = c(0.75, 0.75, 0.50, 1.00, 0.10, 0.10, 1.00,
                0.30, 0.50, 0.30, 0.50, 0.20)
)

collapsed <- collapsed %>%
  left_join(anchoring_thresholds, by = "common_name_clean") %>%
  mutate(
    # Zero out steps below species-specific threshold
    step_anchored = ifelse(!is.na(step_distance) & step_distance < threshold,
                           0, step_distance)
  )

cat("Example anchoring thresholds:\n")
print(anchoring_thresholds)
```

## 3.4 Sequence-level filters and speed calculation

```{r sequence-speed}
# Compute sequence-level speeds
sequence_speeds <- collapsed %>%
  group_by(sequence_id_use, deployment_id_clean, common_name_clean) %>%
  summarise(
    n_frames = n(),
    n_unique_times = n_distinct(unique_sec),
    duration_secs = max(time_sec) - min(time_sec),
    path_distance = sum(step_anchored, na.rm = TRUE),
    group_size = max(1, n_distinct(detection_idx), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(speed = ifelse(duration_secs > 0, path_distance / duration_secs, NA_real_))

# Apply sequence-level filters
sequence_filtered <- sequence_speeds %>%
  filter(
    n_unique_times >= 2,          # Multi-frame requirement
    duration_secs > 0,            # Non-zero duration
    !is.na(speed),                # Valid speed
    speed > 0,                    # Non-zero speed
    speed < 20,                   # Biological plausibility (<20 m/s)
    group_size == 1               # Single-animal sequences only
  )

cat("Sequences before filtering:", nrow(sequence_speeds), "\n")
cat("Sequences after filtering:", nrow(sequence_filtered), "\n")
```

## 3.5 Tortuosity filter

Sequences where accumulated coordinate noise produces biologically implausible zigzag paths are removed.

```{r tortuosity}
# Compute tortuosity (path distance / straight-line distance)
tortuosity <- collapsed %>%
  group_by(sequence_id_use) %>%
  summarise(
    start_x = first(world_x), start_z = first(world_z),
    end_x = last(world_x), end_z = last(world_z),
    .groups = "drop"
  ) %>%
  mutate(
    straight_line = sqrt((end_x - start_x)^2 + (end_z - start_z)^2),
    straight_line = ifelse(straight_line == 0, NA_real_, straight_line)
  )

sequence_filtered <- sequence_filtered %>%
  left_join(tortuosity %>% select(sequence_id_use, straight_line),
            by = "sequence_id_use") %>%
  mutate(
    tortuosity = ifelse(!is.na(straight_line) & straight_line > 0,
                        path_distance / straight_line, NA_real_)
  ) %>%
  left_join(anchoring_thresholds, by = "common_name_clean") %>%
  mutate(
    tort_threshold = case_when(
      common_name_clean %in% c("Red Squirrel", "Snowshoe Hare") ~ 5,
      common_name_clean == "Elk" ~ 4,
      TRUE ~ 3
    )
  ) %>%
  filter(is.na(tortuosity) | tortuosity <= tort_threshold)

cat("Sequences after tortuosity filter:", nrow(sequence_filtered), "\n")
```

## 3.6 SBD speed estimation

The harmonic mean corrects for size-biased sampling, where faster animals are over-represented in the sample.

```{r sbd-speed}
# SBD speed = harmonic mean
sbd_by_species <- sequence_filtered %>%
  group_by(common_name_clean) %>%
  filter(n() >= 10) %>%  # Minimum 10 sequences
  summarise(
    n = n(),
    sbd_speed = n / sum(1 / speed),    # Harmonic mean
    arithmetic_mean = mean(speed),
    .groups = "drop"
  )

# Bootstrap 95% CIs
set.seed(42)
boot_sbd <- function(speeds, n_boot = 1000) {
  boots <- replicate(n_boot, {
    s <- sample(speeds, length(speeds), replace = TRUE)
    length(s) / sum(1 / s)
  })
  c(lcl = quantile(boots, 0.025), ucl = quantile(boots, 0.975),
    se = sd(boots))
}

sbd_results <- sequence_filtered %>%
  group_by(common_name_clean) %>%
  filter(n() >= 10) %>%
  group_modify(~ {
    ci <- boot_sbd(.x$speed)
    tibble(
      n = nrow(.x),
      sbd_speed = nrow(.x) / sum(1 / .x$speed),
      se = ci["se"],
      lcl = ci["lcl.2.5%"],
      ucl = ci["ucl.97.5%"]
    )
  }) %>%
  ungroup()

cat("SBD speed estimates:\n")
sbd_results %>%
  mutate(across(where(is.numeric) & !matches("^n$"), ~round(., 3))) %>%
  knitr::kable(col.names = c("Species", "n", "SBD Speed (m/s)", "SE", "LCL", "UCL"))
```

> **Note:** Only species with ≥10 movement sequences receive speed estimates. Species with fewer sequences (e.g., rare or low-detection species) will show as NA in the final summary. With data from additional arrays or longer deployment periods, more species would meet this threshold. For applied density estimation, we recommend ≥30 sequences per species.
```

# 4. Activity level estimation

Activity level represents the proportion of the 24-hour cycle during which animals are active (Rowcliffe et al. 2014). We use the `activity` package to fit a circular kernel density to detection timestamps.

```{r activity}
# Convert detection times to radians (one per sequence, not per frame)
activity_data <- data_clean %>%
  group_by(sequence_id_use, common_name_clean) %>%
  slice_min(timestamp_clean, n = 1) %>%
  ungroup() %>%
  mutate(time_rad = (hour / 24) * 2 * pi)

# Estimate activity for species with sufficient detections
activity_results <- activity_data %>%
  group_by(common_name_clean) %>%
  filter(n() >= 40) %>%
  group_modify(~ {
    fit <- fitact(.x$time_rad, reps = 100, sample = "data")
    tibble(
      n = nrow(.x),
      activity = fit@act[1],
      se = fit@act[2],
      lcl = fit@act[3],
      ucl = fit@act[4]
    )
  }) %>%
  ungroup()

cat("Activity level estimates:\n")
activity_results %>%
  mutate(across(where(is.numeric) & !matches("^n$"), ~round(., 3))) %>%
  knitr::kable(col.names = c("Species", "n", "Activity", "SE", "LCL", "UCL"))
```

> **Note:** Activity estimation requires ≥40 independent detections for stable kernel density fitting. Species below this threshold will appear as NA in the final summary. Activity estimates can also vary across arrays and seasons, so site-specific estimates are preferable when sample sizes allow.
```

## Activity plots

```{r activity-plots, fig.width=10, fig.height=8}
plot_species <- activity_results$common_name_clean

if (length(plot_species) > 0) {
  par(mfrow = c(ceiling(length(plot_species) / 2), 2), mar = c(4, 4, 3, 1))
  
  for (sp in plot_species) {
    sp_times <- activity_data %>%
      filter(common_name_clean == sp) %>%
      pull(time_rad)
    
    fit <- fitact(sp_times, reps = 100, sample = "data")
    plot(fit, main = sp, ylab = "Detection frequency", xlab = "Time of day",
         tline = list(col = "darkgreen", lwd = 2))
  }
}
```

# 5. Staying time estimation

Staying time is the total duration an animal is present within the camera's detection zone. We construct a 1-second timeline for each sequence and interpolate gaps using edge detection logic.

## 5.1 Build per-sequence timelines

```{r staying-time}
# Get bounding box edge information
# Edge = bbox within outer 1/6 of frame width
EDGE_FRAC <- 1/6

staying_data <- data_clean %>%
  mutate(
    at_left_edge = x1 < (IMG_WIDTH * EDGE_FRAC),
    at_right_edge = x2 > (IMG_WIDTH * (1 - EDGE_FRAC)),
    at_edge = at_left_edge | at_right_edge,
    edge_side = case_when(
      at_left_edge ~ "left",
      at_right_edge ~ "right",
      TRUE ~ "none"
    )
  )

# Compute staying time per sequence
compute_staying_time <- function(seq_data) {
  
  times <- sort(unique(floor(as.numeric(
    difftime(seq_data$timestamp_clean, min(seq_data$timestamp_clean), units = "secs")
  ))))
  
  if (length(times) <= 1) return(tibble(st_g = 1, st_i = 1))
  
  # Build 1-second timeline
  full_timeline <- 0:max(times)
  n_detected <- rep(0, length(full_timeline))
  
  # Fill in observed seconds
  for (t in times) {
    idx <- which(full_timeline == t)
    n_at_t <- seq_data %>%
      filter(floor(as.numeric(difftime(timestamp_clean, min(seq_data$timestamp_clean),
                                        units = "secs"))) == t) %>%
      nrow()
    n_detected[idx] <- max(1, n_at_t)
  }
  
  # Simple gap interpolation: fill short gaps (<=10 sec) between detections
  for (i in seq_along(full_timeline)) {
    if (n_detected[i] == 0) {
      # Find nearest detection before and after
      before <- max(which(n_detected[1:i] > 0), na.rm = TRUE)
      after_idx <- which(n_detected[i:length(n_detected)] > 0)
      after <- ifelse(length(after_idx) > 0, i + min(after_idx) - 1, NA)
      
      if (!is.na(before) && !is.na(after)) {
        gap_length <- after - before - 1
        if (gap_length <= 10) {
          # Check edge conditions
          before_edge <- seq_data %>%
            filter(floor(as.numeric(difftime(timestamp_clean,
                                             min(seq_data$timestamp_clean),
                                             units = "secs"))) == full_timeline[before]) %>%
            pull(edge_side) %>% first()
          after_edge <- seq_data %>%
            filter(floor(as.numeric(difftime(timestamp_clean,
                                             min(seq_data$timestamp_clean),
                                             units = "secs"))) == full_timeline[after]) %>%
            pull(edge_side) %>% first()
          
          # Same edge = animal left and returned, set to absent
          if (!is.na(before_edge) && !is.na(after_edge) &&
              before_edge == after_edge && before_edge != "none") {
            n_detected[i] <- 0
          } else {
            # Otherwise interpolate as present
            n_detected[i] <- round(mean(c(n_detected[before], n_detected[after])))
            n_detected[i] <- max(1, n_detected[i])
          }
        }
      }
    }
  }
  
  st_g <- sum(n_detected > 0)  # Group staying time (presence-seconds)
  st_i <- sum(n_detected)       # Individual staying time (animal-seconds)
  
  tibble(st_g = st_g, st_i = st_i)
}

# Apply to all sequences
staying_results <- staying_data %>%
  group_by(sequence_id_use, deployment_id_clean, common_name_clean) %>%
  group_modify(~ compute_staying_time(.x)) %>%
  ungroup()

# Summarise by species
staying_by_species <- staying_results %>%
  group_by(common_name_clean) %>%
  summarise(
    n = n(),
    mean_st_g = mean(st_g),
    median_st_g = median(st_g),
    mean_st_i = mean(st_i),
    .groups = "drop"
  )

cat("Staying time estimates:\n")
staying_by_species %>%
  mutate(across(where(is.numeric) & !matches("^n$"), ~round(., 1))) %>%
  knitr::kable(col.names = c("Species", "n", "Mean ST_g (s)", "Median ST_g (s)", "Mean ST_i (s)"))
```

# 6. Effective detection distance

We estimate deployment-specific effective detection distances using a joint multi-species hierarchical model. The detection function scale parameter is decomposed into a shared deployment-level effect and species-specific offsets.

## 6.1 Prepare detection histograms

We use only the trigger picture (first frame of each sequence) as the detection distance, ensuring independence among observations.

```{r edd-prep}
# One distance per sequence (trigger picture)
trigger_dists <- data_clean %>%
  group_by(sequence_id_use, deployment_id_clean, common_name_clean) %>%
  slice_min(timestamp_clean, n = 1) %>%
  ungroup() %>%
  filter(world_z > 0, world_z <= 26)

cat("Trigger pictures:", nrow(trigger_dists), "\n")

# Build detection histograms
BIN_WIDTH <- 3
MAX_DIST <- 27
breaks <- seq(0, MAX_DIST, by = BIN_WIDTH)
n_bins <- length(breaks) - 1

# Create species-deployment combinations
sp_dep <- trigger_dists %>%
  group_by(common_name_clean, deployment_id_clean) %>%
  summarise(n_det = n(), .groups = "drop")

cat("Species-deployment combinations:", nrow(sp_dep), "\n")
cat("Species represented:", n_distinct(sp_dep$common_name_clean), "\n")
```

## 6.2 Build data for NIMBLE

```{r edd-nimble-data}
# Species and deployment indices
species_list <- sort(unique(trigger_dists$common_name_clean))
deploy_list <- sort(unique(trigger_dists$deployment_id_clean))
K <- length(species_list)
I <- length(deploy_list)

# Build observation-level data
obs_list <- list()
obs_idx <- 0

for (s in seq_along(species_list)) {
  for (d in seq_along(deploy_list)) {
    dets <- trigger_dists %>%
      filter(common_name_clean == species_list[s],
             deployment_id_clean == deploy_list[d])
    
    if (nrow(dets) == 0) next
    
    obs_idx <- obs_idx + 1
    
    # Bin the detections
    hist_counts <- hist(dets$world_z, breaks = breaks, plot = FALSE)$counts
    
    obs_list[[obs_idx]] <- list(
      species_idx = s,
      deploy_idx = d,
      counts = hist_counts,
      n_det = sum(hist_counts)
    )
  }
}

M <- length(obs_list)
cat("Total observations (species x deployment):", M, "\n")

# Construct data matrices
L <- matrix(0, nrow = M, ncol = n_bins)
species_idx <- integer(M)
deploy_idx <- integer(M)
N_obs <- integer(M)

for (m in seq_along(obs_list)) {
  L[m, ] <- obs_list[[m]]$counts
  species_idx[m] <- obs_list[[m]]$species_idx
  deploy_idx[m] <- obs_list[[m]]$deploy_idx
  N_obs[m] <- obs_list[[m]]$n_det
}

# Bin midpoints and area weights
bin_mids <- breaks[-length(breaks)] + BIN_WIDTH / 2
bin_areas <- (breaks[-1]^2 - breaks[-length(breaks)]^2) / MAX_DIST^2

cat("Bins:", n_bins, "| Bin width:", BIN_WIDTH, "m | Max distance:", MAX_DIST, "m\n")
cat("Species:", K, "| Deployments:", I, "| Observations:", M, "\n")
```

## 6.3 Define and run the NIMBLE model

```{r edd-model, results='hide'}
edd_code <- nimbleCode({
  # Priors
  gamma0 ~ dnorm(0, sd = 10)
  tau_alpha ~ dgamma(0.01, 0.01)
  sd_alpha <- 1 / sqrt(tau_alpha)
  tau_delta ~ dgamma(0.01, 0.01)
  sd_delta <- 1 / sqrt(tau_delta)
  
  # Species offsets (K-1 free, Kth derived for sum-to-zero)
  for (k in 1:(K-1)) {
    delta_free[k] ~ dnorm(0, sd = sd_delta)
  }
  # Sum-to-zero constraint
  delta[1:(K-1)] <- delta_free[1:(K-1)]
  delta[K] <- -sum(delta_free[1:(K-1)])
  
  # Deployment random effects
  for (i in 1:I) {
    alpha[i] ~ dnorm(gamma0, sd = sd_alpha)
  }
  
  # Likelihood
  for (m in 1:M) {
    # Detection scale for this species-deployment combination
    log_sigma[m] <- alpha[deploy_idx[m]] + delta[species_idx[m]]
    sigma[m] <- exp(log_sigma[m])
    
    # Half-normal detection probabilities per bin
    for (j in 1:n_bins) {
      p[m, j] <- exp(-bin_mids[j]^2 / (2 * sigma[m]^2))
      pi_raw[m, j] <- bin_areas[j] * p[m, j]
    }
    
    # Normalize to proper probabilities
    pi_sum[m] <- sum(pi_raw[m, 1:n_bins])
    for (j in 1:n_bins) {
      pi_norm[m, j] <- pi_raw[m, j] / pi_sum[m]
    }
    
    # Multinomial likelihood
    L[m, 1:n_bins] ~ dmulti(pi_norm[m, 1:n_bins], N_obs[m])
  }
})

# Data and constants
edd_constants <- list(
  K = K, I = I, M = M, n_bins = n_bins,
  species_idx = species_idx,
  deploy_idx = deploy_idx,
  N_obs = N_obs,
  bin_mids = bin_mids,
  bin_areas = bin_areas
)

edd_data <- list(L = L)

# Initial values
edd_inits <- list(
  gamma0 = log(5),
  tau_alpha = 1,
  tau_delta = 10,
  delta_free = rep(0, K - 1),
  alpha = rep(log(5), I)
)

# Build and run
model <- nimbleModel(edd_code, constants = edd_constants,
                     data = edd_data, inits = edd_inits)
compiled <- compileNimble(model)
mcmc_conf <- configureMCMC(model,
                            monitors = c("gamma0", "alpha", "delta",
                                         "sd_alpha", "sd_delta",
                                         "tau_alpha", "tau_delta"))
mcmc <- buildMCMC(mcmc_conf)
compiled_mcmc <- compileNimble(mcmc, project = model)

# Run MCMC
samples <- runMCMC(compiled_mcmc,
                   niter = 10000,
                   nburnin = 2000,
                   thin = 5,
                   nchains = 2,
                   setSeed = TRUE)
```

## 6.4 Extract EDD estimates

```{r edd-extract}
# Combine chains
all_samples <- do.call(rbind, samples)

# Extract deployment-level alphas and species deltas
alpha_cols <- grep("^alpha\\[", colnames(all_samples), value = TRUE)
delta_cols <- grep("^delta\\[", colnames(all_samples), value = TRUE)

alpha_post <- all_samples[, alpha_cols]
delta_post <- all_samples[, delta_cols]

# Compute EDD for each species-deployment combination
# EDD = B * sqrt(sum(p_j * a_j)) where p_j is detection prob at bin j
edd_estimates <- list()

for (m in seq_len(M)) {
  s <- species_idx[m]
  d <- deploy_idx[m]
  
  # Posterior of log_sigma for this observation
  log_sigma_post <- alpha_post[, d] + delta_post[, s]
  sigma_post <- exp(log_sigma_post)
  
  # EDD = sigma * sqrt(pi/2) for half-normal
  edd_post <- sigma_post * sqrt(pi / 2)
  
  edd_estimates[[m]] <- tibble(
    species = species_list[s],
    deployment = deploy_list[d],
    EDD_mean = mean(edd_post),
    EDD_sd = sd(edd_post),
    EDD_lcl = quantile(edd_post, 0.025),
    EDD_ucl = quantile(edd_post, 0.975)
  )
}

edd_df <- bind_rows(edd_estimates)

# Summary by species
edd_by_species <- edd_df %>%
  group_by(species) %>%
  summarise(
    n_deployments = n(),
    mean_EDD = mean(EDD_mean),
    sd_EDD = sd(EDD_mean),
    .groups = "drop"
  )

cat("EDD estimates by species:\n")
edd_by_species %>%
  mutate(across(where(is.numeric) & !matches("n_"), ~round(., 2))) %>%
  knitr::kable(col.names = c("Species", "Deployments", "Mean EDD (m)", "SD"))
```

## 6.5 Model diagnostics

```{r edd-diagnostics}
# Check convergence for key parameters
cat("Convergence diagnostics:\n")
cat("  gamma0 Rhat:", round(MCMCsummary(samples, params = "gamma0")$Rhat, 3), "\n")
cat("  sd_alpha Rhat:", round(MCMCsummary(samples, params = "sd_alpha")$Rhat, 3), "\n")

# Species offsets
delta_summary <- MCMCsummary(samples, params = "delta")
cat("\nSpecies offsets (delta):\n")
delta_table <- tibble(
  species = species_list,
  delta_mean = delta_summary$mean,
  delta_lcl = delta_summary$`2.5%`,
  delta_ucl = delta_summary$`97.5%`
)
delta_table %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  knitr::kable()
```

# 7. Summary: all parameters

```{r summary-table}
# Combine all parameter estimates
final_params <- sbd_results %>%
  select(species = common_name_clean, n_speed = n,
         sbd_speed, speed_se = se) %>%
  full_join(
    activity_results %>%
      select(species = common_name_clean, n_activity = n,
             activity, activity_se = se),
    by = "species"
  ) %>%
  full_join(
    staying_by_species %>%
      select(species = common_name_clean, n_staying = n,
             mean_st_g),
    by = "species"
  ) %>%
  full_join(
    edd_by_species %>% select(species, n_edd = n_deployments, mean_EDD),
    by = "species"
  ) %>%
  mutate(
    day_range_km = sbd_speed * activity * 86400 / 1000
  )

cat("=== COMPLETE PARAMETER ESTIMATES ===\n\n")
final_params %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  knitr::kable(col.names = c("Species", "n_speed", "SBD Speed (m/s)", "Speed SE",
                              "n_activity", "Activity", "Activity SE",
                              "n_staying", "ST_g (s)",
                              "n_EDD", "EDD (m)", "Day Range (km/d)"))
```

These parameter estimates are direct inputs for density models:

- **REM** uses: speed, activity (as day range), EDD, camera angle
- **REST** uses: staying time, detection area (from EDD and camera angle)  
- **CTDS** uses: EDD, staying time, camera angle

> **Note:** NA values in the summary table indicate species that did not meet minimum sample size requirements for speed (≥10 sequences) or activity (≥40 detections) estimation from this single array. The EDD model and staying time estimation have lower data requirements and typically produce estimates for all detected species. In a full analysis using data from multiple arrays (e.g., the complete study with 181 deployments), most species would have complete parameter sets. Day range can only be calculated when both speed and activity are available.

# References

- Howe EJ et al. 2017. Distance sampling with camera traps. *Methods in Ecology and Evolution* 8: 1558-1565.
- Nakashima Y et al. 2018. Estimating animal density without individual recognition. *Journal of Applied Ecology* 55: 735-744.
- Rowcliffe JM et al. 2008. Estimating animal density using camera traps without the need for individual recognition. *Journal of Applied Ecology* 45: 1228-1236.
- Rowcliffe JM et al. 2014. Quantifying levels of animal activity using camera trap data. *Methods in Ecology and Evolution* 5: 1170-1179.
- Rowcliffe JM et al. 2016. Wildlife speed cameras: measuring animal travel speed and day range. *Remote Sensing in Ecology and Conservation* 2: 84-94.

# Session info

```{r session}
sessionInfo()
```
